\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-03 NAR_DRAFT.tex

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple statistical model and a well-tested set of heuristic filters.
However, in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
The algorithm has low computational overhead, and in principle there should be no bar to simply incorporating it into existing variant calling algorithms.

>>>>

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.
